---
title: "Data622 - Group2 - Homework4"
author: "Zachary Palmore, Kevin Potter, Amit Kapoor, Adam Gersowitz, Paul Perez"
date: "10/21/2021"
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
  html_document:
    fig_width: 15
    highlight: pygments
    number_sections: no
    theme: flatly
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width = 10)
```


# Overview


# Approach


```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# Libraries
set.seed(622)
```

# Data Exploration
## Data Characteristics


```{r data}
# read data

```


## Data summary


```{r loan_data_summary}

```



# Data Preparation

## Handling missing values

```{r}
# plot missing values
plot_missing(loan_data)
```



```{r cat-missing}
# Filter out 

```



```{r num-missing}
# impute numeric predictors using mice

```


```{r}
dim(loan_data)
```


## Preprocess using transformation


```{r transform-train}

```


## Training and Test Partition

In this step for data preparation we will partition the training dataset in training and validation sets using `createDataPartition` method from `caret` package. We will reserve 75% for training and rest 25% for validation purpose.

```{r partition}
set.seed(622)
partition <- createDataPartition(loan_data$Loan_Status, p=0.75, list = FALSE)

training <- loan_data[partition,]
testing <- loan_data[-partition,]

# training/validation partition for independent variables
#X.train <- ld.clean[partition, ] %>% dplyr::select(-Loan_Status)
#X.test <- ld.clean[-partition, ] %>% dplyr::select(-Loan_Status)

# training/validation partition for dependent variable Loan_Status
#y.train <- ld.clean$Loan_Status[partition]
#y.test <- ld.clean$Loan_Status[-partition]
```


# Build Models

## Linear Discriminant Analysis (LDA)

## K-nearest neighbor (KNN)

## Decision Trees


## Random Forests


# Model Performance



# Conclusion


# References


https://www.r-bloggers.com/2018/07/prop-table/

https://www.datacamp.com/community/tutorials/decision-trees-R

https://stats.stackexchange.com/questions/82162/cohens-kappa-in-plain-english

# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

















