---
title: "Data622 - Group2 - Homework4"
author: "Zachary Palmore, Kevin Potter, Amit Kapoor, Adam Gersowitz, Paul Perez"
date: "10/21/2021"
output:
  html_document:
    fig_width: 15
    highlight: pygments
    number_sections: no
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    latex_engine: xelatex
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width = 10)
```


# Overview
In this project, we will analyze a real-life mental health dataset to provide context around suicide prediction given the variety of unidentifiable demographic data.

# Approach
We will first perform exploratory data analysis on dataset to later inform our modeling approaches for Clustering, Principal Compnent Analysis, Gradient Boosting, and Support Vector Machines.

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# Libraries

library(summarytools)
library(tidyverse)
library(DataExplorer)
library(reshape2)
library(mice)
library(caret)
library(MASS)
library(e1071)
library(tree)
library(corrplot)
library(kableExtra)
library(htmltools)
library(readxl)
library(psych)

set.seed(622)

set.seed(622)
```

# Data Exploration

Columns | Variable | Description  
---|---|-----  
C |  Sex | Male-1, Female-2  
D | Race | White-1, African American-2, Hispanic-3, Asian-4, Native American-5, Other or missing data -6  
E - W | ADHD self-report scale |  Never-0, rarely-1, sometimes-2, often-3, very often-4  
X - AM | Mood disorder questions |  No-0, yes-1; question 3: no problem-0, minor-1, moderate-2, serious-3  
AN - AS | Individual substances misuse |   no use-0, use-1, abuse-2, dependence-3   
AT | Court Order |   No-0, Yes-1  
AU | Education |  1-12 grade, 13+ college  
AV | History of Violence |  No-0, Yes-1  
AW | Disorderly Conduct |  No-0, Yes-1  
AX | Suicide attempt |  No-0, Yes-1  
AY | Abuse Hx |  No-0, Physical (P)-1, Sexual (S)-2, Emotional (E)-3, P&S-4, P&E-5, S&E-6, P&S&E-7  
AZ | Non-substance-related Dx |  0 - none; 1 - one; 2 - More than one  
BA | Substance-related Dx |  0 - none; 1 - one Substance-related; 2 - two; 3 - three or more  
BB | Psychiatric Meds |  0 - none; 1 - one psychotropic med; 2 - more than one psychotropic med  


## Data Characteristics


```{r data}
# read data
adhd_data <- read_excel("ADHD_data.xlsx", sheet = "Data") %>% na_if("") %>% dplyr::select(-1)
#columns <- list(dimnames(adhd_data)[2])
#df <- adhd_data[,2:53]
adhd_data[,2:53] <- lapply(adhd_data[,2:53], factor)
```


```{r}
dim(adhd_data)
```

```{r}
adhd_data[,c(23:37)]
```


```{r, cat-bar, fig.length =20, fig.width=10}
# select categorical columns
cat_cols <- dimnames(adhd_data[,2:53])[[2]]
adhd_fact <-  adhd_data[cat_cols]
# long format
adhd_factm <- melt(adhd_fact, measure.vars = cat_cols, variable.name = 'metric', value.name = 'value')
# plot categorical columns
ggplot(adhd_factm, aes(x = value)) + 
  geom_bar(aes(fill = metric)) + 
  facet_wrap( ~ metric, nrow = 5L, scales = 'free') + coord_flip() + 
  theme(legend.position = "none")
```

## Data summary


```{r adhd_data_summary}
dfSummary(adhd_data, style = 'grid', graph.col = FALSE)
```

## Coorelation

Next we will see the correlation among ADHD questions and MD questions. As we can deduce from below 2 correlation plots, ADHD questions are highly correlated and MD questions comparatively shows moderate correlation.

```{r corr-adhds}
adhds <- sapply(adhd_data[,c(4:21)], as.numeric) %>% cor()
corrplot::corrplot(adhds, method="number")
```

```{r corr-mds}
mds <- sapply(adhd_data[,c(23:37)], as.numeric) %>% cor()
corrplot::corrplot(mds, method="number")
```



# Data Preparation

## Factor Analysis

Like PCA, Factor Analysis too, reduces larger number of variables into smaller number of variables, called latent variables.It is used to identify underlying factors that explain the correlation among set of variables. Factor analysis is a great tool for treating multivariate questionnaire studies.

For ADHD questions, test of the hypothesis that 3 factors are sufficient. The chi square statistic is 197.3 on 102 degrees of freedom. The p-value is 0.0000000476. We have used regression factor scores here as they predict the location of each individual on the factor.



```{r adhd-ques-fa}
adhd_ques_fa <- factanal(sapply(adhd_data[,c(4:21)], as.numeric), 
                         factors = 3, 
                         rotation = "promax", 
                         scores = "regression")
adhd_ques_fa
```



```{r adhd-fa-diag}
fa.diagram(adhd_ques_fa$loadings)
```


For MD questions we could see that 1st MD question has multiple sub questions as compared to 2nd and 3rd question. Now for these set of MD questions too, we will apply similar factor analysis as of ADHD questions. Test of the hypothesis that 3 factors are sufficient. The chi square statistic is 88.82 on 63 degrees of freedom. The p-value is 0.0178.



```{r md-ques-fa}
md_ques_fa <- factanal(sapply(adhd_data[,c(23:37)], as.numeric), 
                         factors = 3, 
                         rotation = "promax", 
                         scores = "regression")
md_ques_fa
```



```{r mdfa-diag}
fa.diagram(md_ques_fa$loadings)
```

In the next step we will remove all ADHD Question columns, ADHD Total, MD questions columns and MD TOTAL columns. Then we will add the new factors found above for ADHD and MD questions.


```{r}
# ADHD question scores dataframe
adhd_ques_fa <- as.data.frame(adhd_ques_fa$scores) 
names(adhd_ques_fa) <- c('ADHD_FACT1','ADHD_FACT2','ADHD_FACT3')

# MD questions scores dataframe
md_ques_fa <- as.data.frame(md_ques_fa$scores)
names(md_ques_fa) <- c('MD_FACT1','MD_FACT2','MD_FACT3')

# remove ADHD and MD columns
adhd_newdata <- adhd_data %>% dplyr::select(-c(starts_with('ADHD Q'), starts_with('MD Q')))

# Add new factor columns created
adhd_newdata <- cbind(adhd_newdata, adhd_ques_fa, md_ques_fa)
```

Here is glimpse of new set of data.

```{r}
head(adhd_newdata)
```




## Handling missing values


```{r}
# plot missing values
plot_missing(adhd_newdata)
```

We can see from this chart that `Psych meds.` contributes to 67.43% of missing data which is maximum among all missing data in other columns. We will remove this column before imputation. We then impute values using MICE (Multivariate Imputation by Chained Equations) for columns having missing values.

```{r}
# rename columns to apply mice
adhd_newdata <- adhd_newdata %>% 
  rename('ADHD_Total'='ADHD Total',
         'MD_Total'='MD TOTAL',
         'Sedative_hypnotics'='Sedative-hypnotics', 
         'Court_order' = 'Court order', 
         'Hx_of_Violence'='Hx of Violence', 
         'Disorderly_Conduct'='Disorderly Conduct', 
         'Non_subst_Dx'='Non-subst Dx',
         'Subst_Dx'='Subst Dx', 
         'Psych_meds'='Psych meds.') %>% 
  dplyr::select(-Psych_meds)
```



```{r impute}
# select columns with non missing values
temp <- adhd_newdata %>% dplyr::select(c(starts_with('ADHD_'), starts_with('MD_'), 'Race', 'Sex', 'Age'))

# impute predictors using mice
adhd_impute <- adhd_newdata %>% dplyr::select(-c(starts_with('ADHD_'), starts_with('MD_'), 'Race', 'Sex', 'Age'))
adhd_impute <- complete(mice(data=adhd_impute, print=FALSE))
summary(adhd_impute)
```



```{r}
# Merged the imputed dataframe with temp
adhd_newdata <- cbind(adhd_impute, temp)
head(adhd_newdata)
```



```{r cat-missing}
# Filter out 
#adhd_data <- adhd_data %>% filter(!is.na(Alcohol) &
#                                  !is.na(THC) &
#                                  !is.na(Cocaine) &
#                                  !is.na(Stimulants) &
#                                  !is.na(`Sedative-hypnotics`) &
#                                  !is.na(Opioids) &
#                                  !is.na(`Court order`) &
#                                  !is.na(Education) &
#                                  !is.na(`Hx of Violence`) &
#                                  !is.na(`Disorderly Conduct`) &
#                                  !is.na(Suicide) &
#                                  !is.na(Abuse) &
#                                  !is.na(`Non-subst Dx`) &
#                                  !is.na(`Subst Dx`) &
#                                  !is.na(`Psych meds.`))
```



```{r num-missing}
# impute numeric predictors using mice
#adhd_data <- complete(mice(data=adhd_data[,:53], method="pmm", print=FALSE))
```




## Preprocess using transformation


```{r transform}
set.seed(622)
adhd_transform <- adhd_mice %>% 
  dplyr::select(c("Age","ADHD_FACT1","ADHD_FACT2","ADHD_FACT3",,"MD_FACT1","MD_FACT2","MD_FACT3")) %>%
  preProcess(method = c("center","scale")) %>% 
  predict(adhd_mice)
```


## Training and Test Partition

In this step for data preparation we will partition the training dataset in training and validation sets using `createDataPartition` method from `caret` package. We will reserve 75% for training and rest 25% for validation purpose.

```{r partition}
set.seed(622)
partition <- createDataPartition(adhd_data$Suicide, p=0.75, list = FALSE)
training <- adhd_data[partition,]
testing <- adhd_data[-partition,]
# training/validation partition for independent variables
#X.train <- ld.clean[partition, ] %>% dplyr::select(-Loan_Status)
#X.test <- ld.clean[-partition, ] %>% dplyr::select(-Loan_Status)
# training/validation partition for dependent variable Loan_Status
#y.train <- ld.clean$Loan_Status[partition]
#y.test <- ld.clean$Loan_Status[-partition]
```

# Clustering Models
## K-means
## Hierarchical

# Principal Component Analysis
## Individual Substance Misuse

# Gradient Boosting: Suicide

# Support Vector

# Build Models

## Linear Discriminant Analysis (LDA)

## Clustering Method

We use K-nearest neighbor (KNN) to identify clusters of patients that share similar patterns that could help us predict our target variable. 

```{r cluster method knn}
set.seed(622)
mode <- function(x){
  levels <- unique(x)
  indicies <- tabulate(match(x, levels))
  levels[which.max(indicies)]
}
# Clean up training data
training_factors <- training %>% 
  dplyr::select(-Age, -`ADHD Total`, `MD TOTAL`) 
training_factors <- data.frame(lapply(training_factors, as.factor))
train_knn <- training_factors %>% 
  mutate(across(everything(), ~replace_na(., mode(.))))
mode(train_knn$Psych.meds.) 
train_knn$Psych.meds.[which(is.na(train_knn$Psych.meds.))] <- 0
sum(is.na(train_knn$Psych.meds.))

# Clean up testing data
testing_factors <- testing %>% 
  dplyr::select(-Age, -`ADHD Total`, `MD TOTAL`) 
testing_factors <- data.frame(lapply(testing_factors, as.factor))
test_knn <- testing_factors %>% 
  mutate(across(everything(), ~replace_na(., mode(.))))
mode(test_knn$Psych.meds.) 
test_knn$Psych.meds.[which(is.na(test_knn$Psych.meds.))] <- 0
sum(is.na(test_knn$Psych.meds.))

# Train KNN model
train.knn <- (train_knn[, names(train_knn) != "Suicide"])
prep <- preProcess(x = train.knn, method = c("center", "scale"))
cl <- trainControl(method="repeatedcv", repeats = 5) 
knn_model <- train(Suicide ~ ., data = train_knn, 
                method = "knn", 
                trControl = cl, 
                preProcess = c("center","scale"), 
                tuneLength = 20)
knn_model 
```


```{r}
# Evaluate Model
plot(knn_model)
knn_predict <- predict(knn_model, newdata = test_knn)
mean(knn_predict == test_knn$Suicide) # accuracy
conf.mat.knn <- confusionMatrix(knn_predict, test_knn$Suicide)
accuracy <- round(conf.mat.knn$overall[[1]], 3)*100
conf.mat.knn
```

Our KNN model accuracy comes out to `r round(conf.mat.knn$overall[[1]], 3)*100`%


## Decision Trees


## Random Forests


# Model Performance



# Conclusion


# References

https://towardsdatascience.com/what-is-the-difference-between-pca-and-factor-analysis-5362ef6fa6f9

https://scholarworks.umass.edu/cgi/viewcontent.cgi?article=1226&context=pare

# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

















