---
title: "Data622 - Group2 - Homework3"
author: "Zachary Palmore, Kevin Potter, Amit Kapoor"
date: "10/2/2021"
output:
  html_document:
    fig_width: 15
    highlight: pygments
    number_sections: no
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    latex_engine: xelatex
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.align="center", fig.width = 10)
```


# Overview
In this project, the dataset used, is for Loan approval where the prediction will be done for Loan approval status using Linear Discriminant Analysis (LDA), K-nearest neighbor (KNN), Decision Trees and Random Forest models.

# R packages
We will use `r` for data modeling. All packages used for data exploration, visualization, preparation and modeling are listed in Code Appendix.

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# Libraries

library(summarytools)
library(tidyverse)

set.seed(622)
```

# Data Exploration
Below is the description of the variables of interest in the data set.

|VARIABLE NAME|DESCRIPTION|
|--|----|
|Loan_ID|Unique Loan ID|
|Gender|Male/ Female|
|Married|Applicant married (Y/N)|
|Dependents|Number of dependents|
|Education|Applicant Education (Graduate/ Undergraduate)|
|Self_Employed|Self employed (Y/N)|
|ApplicantIncome|Applicant income|
|CoapplicantIncome|Coapplicant income|
|LoanAmount|Loan amount in thousands|
|Loan_Amount_Term|Term of loan in months|
|Credit_History|credit history meets guidelines|
|Property_Area|Urban/ Semi Urban/ Rural|
|Loan_Status|Loan approved (Y/N)|


```{r data}
# read data, change blank to NA and and remove loan_id
loan_data <- read.csv('https://raw.githubusercontent.com/amit-kapoor/Data622Group2/main/Loan_approval.csv') %>% 
  na_if("") %>%
  select(-1)

# categorical columns as factors
loan_data <- loan_data %>% 
  mutate(Gender=as.factor(Gender),
         Married=as.factor(Married),
         Dependents=as.factor(Dependents),
         Education=as.factor(Education),
         Self_Employed=as.factor(Self_Employed),
         Property_Area=as.factor(Property_Area),
         Loan_Status=as.factor(Loan_Status))

```


## Data summary

Below is summary of loan approval dataset.

```{r loan_data_summary}
dfSummary(loan_data, style = 'grid', graph.col = FALSE)
```

* There are 7 columns having missing values.
* The proportion of values for few columns shows significant differences i.e. Gender (more males), Married( more married), Credit_History (more having credit history).


# Data Preparation

## Handling missing values

## Training and Test Partition


# Build Models

## Linear Discriminant Analysis (LDA)


## K-nearest neighbor (KNN)

## Decision Trees

## Random Forests



# Model performance



# Conclusion

# References

# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


















